Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-01-14 paper/main.tex
Updates

% The standard author block has changed for UAI 2025 to provide
% more space for long author lists and allow for complex affiliations
%
% All author information is authomatically removed by the class for the
% anonymous submission version of your paper, so you can already add your
% information below.
%
% Add authors
\author[1]{\href{mailto:<jj@example.edu>?Subject=Your UAI 2025 paper}{Jane~J.~von~O'L\'opez}{}}
\author[1]{Harry~Q.~Bovik}
\author[1,2]{Further~Coauthor}
\author[3]{Further~Coauthor}
\author[1]{Further~Coauthor}
\author[3]{Further~Coauthor}
\author[3,1]{Further~Coauthor}
% Add affiliations after the authors
\affil[1]{%
    Computer Science Dept.\\
    Cranberry University\\
    Pittsburgh, Pennsylvania, USA
}
\affil[2]{%
    Second Affiliation\\
    Address\\
    …
}
\affil[3]{%
    Another Affiliation\\
    Address\\
    …
  }

\begin{document}

@@

\begin{abstract}
Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. Despite these advancements, the
application of these algorithms in applied domains remains limited, with
researchers often preferring to manually construct DAGs based on their expert
knowledge. This preference stems due to some practical issues with the
automated algorithms such as they can make obvious mistakes, and they output
the Markov Equivalence Class (MEC). Due to these issues, when using these
algorithms, users typically have to rely on expert knowledge to fix the
mistakes and to orient the MEC to a DAG. Given this reliance on manual model
construction, we propose an iterative method to assist researchers in manually
constructing DAGs by combining implied Conditional Independence (CI) based
model testing and utilizing their expert knowledge. Our approach is inspired by
specification search in Structural Equation Models (SEMs), where we utilize a
novel measure of partial association for mixed data to rank violations of
implied Conditional Independences in the model. Using this ranking users can
prioritize the pair of variables which need to be corrected by either adding an
edge between them or removing the edge based on their expertise to specify the
correct orientation. Empirically, we demonstrate that a greedy approach (i.e.,
fixing the worst violation first) achieves performance comparable to automated
algorithms, provided the user can correctly determine the edge orientation in $
1 $ out of $ 3 $ edge additions. Additionally, we show that in the absence of
an expert, a Large Language Model (LLM) can potentially be used to determine
edge orientations. 
\end{abstract}

@@

Despite the many automated algorithms proposed for constructing DAGs, their
adoption in applied research remains limited, with researchers often preferring
to manually construct DAGs based on their domain expertise \citep{Tennant2020,
Petersen2021}. We believe this preference persists due to several issues with
these algorithms when applied in practical settings:

\begin{enumerate}
	\item \textbf{Lack of Trust:} Although most algorithms are
		asymptotically consistent, their finite sample properties are
		not well understood \todo{Is there any citation for this}.
		Moreover, their output can vary significantly depending on the
		chosen algorithm and their hyperparameters. Combined with the
		lack of robust methods to test these algorithms' performance on
		a given dataset, it can be difficult to trust their outputs.
		Consequently, researchers usually have to rely on post-hoc
		modifications to the model based on model testing and their
		domain expertise.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} Multiple
		DAGs can be faithful to a given observational dataset. Thus,
		using only observational data, algorithms can only recover the
		MECs that can contain a combination of directed and undirected
		edges. However, most methods for downstream tasks, such as
		identification or causal effect estimation, assume knowledge of
		a fully oriented DAG. As a result, researchers have to manually
		orient these edges before using them for analysis.
\end{enumerate}

@@

One of the common ways to guide this post-hoc modification of DAGs is to use
Conditional Independence (CI) based model testing \cite{Ankan2023}. As DAGs
imply CI statements that can be tested in the given data using statistical CI
tests, we can verify whether the implied CIs are consistent with the data or
not. If some of implied CIs are not valid in the given dataset, we can modify
the model accordingly. However, this CI based testing has a few drawbacks: 1) CIs
are implied only by missing edges in the model, so we can not test for already
existing edges. 2) In high sample size scenarios, p-values can suggest a
significant association between variables even if they have a very weak
relationship. 3) Number of implied CIs can be large making it difficult to
manually go through each of them; for example,  in the moderately sized alarm
network with $ 37 $ node, there are $287$ implied CIs.

On the other hand, a common method for specification search is using
\emph{modification indices}, which identify potential changes to the model
structure that improve the model fit. These potential changes are then ranked
by the ones that improve the fit the most. This ranking allows researchers to
focus first on the suggestions that can improve the model the most and while
integrating their expertise to choose the most optimal modifications.

Inspired by this ranking approach in modification indices, in this paper we
develop a method to rank the CI tests. We utilize a measure of conditional
association to do this ranking. Additionally, this measure of conditional
association also allows us tackle other issues with CI based testing outlined
in the previous paragraphs.

@@

\begin{figure}
\centering
\begin{subfigure}{0.5 \textwidth}
	\begin{tikzpicture}
	\begin{scope}[xshift=-2.5cm, yshift=0.7cm]
		\tikzstyle{every node}=[align=center, inner sep=1pt]
		\node (sex) at (-0.7, -0.8) {Sex};
		\node (age) at (-0.3, 0) {Age};
		\node (ed) at (-1.8, -1.6) {Education};
		\node (occ) at (1.8, 0) {Occupation};
		\node (hrpw) at (1.3, -1.6) {Hours Per \\ Week};
		\node (income) at (-0.7, -2.8) {Income};

		\draw[->]  (age) to[bend right=20] (ed);
		\draw[->]  (sex) to (ed);
		\draw[->]  (age) to (hrpw);
		\draw[->]  (ed) to (hrpw);
		\draw[->]  (sex) to (hrpw);
		\draw[->]  (ed) to (income);
		\draw[->]  (hrpw) to (income);
		\draw[->]  (occ) to[out=300, in=0, looseness=1.4] (income.east);
		\draw[->]  (sex) to (income);
		\draw[->]  (ed) to[out=90, in=150, looseness=1.3] (occ);
		\draw[->]  (sex) to (occ);	
	\end{scope}
	\end{tikzpicture}
\end{subfigure}
\begin{subfigure}{0.5 \textwidth}
	\begin{tikzpicture}
	\begin{scope}[xshift=-2.5cm, yshift=0.7cm]
		\tikzstyle{every node}=[align=center, inner sep=1pt]
		\node (wclass) at (0, 1) {WorkClass};
		\node (sex) at (0, -0.8) {Sex};
		\node (age) at (-0.3, 0) {Age};
		\node (ed) at (-2, -1.6) {Education};
		\node (occ) at (1.8, 0) {Occupation};
		\node (hrpw) at (1, -2) {Hours Per \\ Week};
		\node (income) at (-0.7, -2.3) {Income};
		\node (marital) at (3, 1) {Marital \\ Status};
		\node (reln) at (-3, 1) {Relationship};
		\node (race) at (-4, 2) {Race};
		\node (country) at (-4, 3) {Native \\ Country};

		\draw[] (ed) to (income);
		\draw[] (marital) to (reln);
		\draw[] (occ) to (sex);
		\draw[->] (reln) to (age);
		\draw[->] (reln) to (marital);
		\draw[] (reln) to (sex);
		\draw[->] (reln) to (hrpw);
		\draw[] (reln) to (income);
		\draw[->] (income) to (ed);
	\end{scope}
	\end{tikzpicture}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
	\begin{tikzpicture}
	\begin{scope}[xshift=-2.5cm, yshift=0.7cm]
		\tikzstyle{every node}=[align=center, inner sep=1pt]
		\node (wclass) at (0, 1) {WorkClass};
		\node (sex) at (0, -0.8) {Sex};
		\node (age) at (-0.3, 0) {Age};
		\node (ed) at (-2, -1.6) {Education};
		\node (occ) at (1.8, 0) {Occupation};
		\node (hrpw) at (1, -2) {Hours Per \\ Week};
		\node (income) at (-0.7, -2.3) {Income};
		\node (marital) at (3, 1) {Marital \\ Status};
		\node (reln) at (-3, 1) {Relationship};
		\node (race) at (-4, 2) {Race};
		\node (country) at (-4, 3) {Native \\ Country};

		\draw[]   (wclass) to (ed);
		\draw[]   (wclass) to (occ);
		\draw[]   (wclass) to (race);
		\draw[]   (wclass) to (income);
		\draw[]   (ed) to (occ);
		\draw[]   (ed) to (country);
		\draw[]   (ed) to (income);
		\draw[]   (marital) to (occ);
		\draw[]   (marital) to (reln);
		\draw[->] (marital) to (sex);
		\draw[]   (marital) to (hrpw);
		\draw[->] (occ) to (sex);
		\draw[]   (occ) to (hrpw);
		\draw[->] (reln) to (sex);
		\draw[]   (reln) to (hrpw);
		\draw[->] (income) to (occ);

	\end{scope}
	\end{tikzpicture}
\end{subfigure}

>>>>

% The standard author block has changed for UAI 2025 to provide
% more space for long author lists and allow for complex affiliations
%
% All author information is authomatically removed by the class for the
% anonymous submission version of your paper, so you can already add your
% information below.
%
% Add authors
\author[1]{\href{mailto:<ankur.ankan@ru.nl>?Subject=Your UAI 2025 paper}{Ankur~Ankan}{}}
\author[1]{Johannes~Textor}

% Add affiliations after the authors
\affil[1]{%
    Institute for Computing and Information Sciences\\
    Radboud University\\
    Nijmegen, The Netherlands
}
\begin{document}

@@

\begin{abstract}
% Causal discovery has received significant attention in the Directed Acyclic
% Graphs (DAGs) literature, leading to the development of numerous automated
% algorithms for learning DAGs from data. Despite these advancements, the
% application of these algorithms in applied domains remains limited, with
% researchers often preferring to manually construct DAGs based on their expert
% knowledge. This preference stems due to some practical issues with these
% algorithms such as making obvious mistakes, and outputting Markov Equivalence
% Class (MEC). As a result of these issues users typically have to rely on post
% hoc modifications to the model based on their expertise. Given this reliance on
% manual model modification and construction, we propose an iterative method to
% assist researchers in manually constructing DAGs by combining implied
% Conditional Independence (CI) based model testing and utilizing their expert
% knowledge. We present a novel measure of partial association for mixed data
% that allows us to rank violations of implied CIs in the model. Using this
% ranking users can prioritize modifications to the model while using their
% expert knowledge to decide orientation of edges. Empirically, we show that
% a greedy approach achieves performance comparable to automated algorithms, provided
% the user can correctly determine the edge orientation in $ 1 $ out of $ 3 $ cases.
% Additionally, we show that in the absence of an expert, a Large Language Model can
% potentially be used for determining edge orientations.

Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. However, their adoption in applied
domains remains limited, as researchers often prefer to construct DAGs manually
based on their expert knowledge. This preference arises due to practical issues
with these automated algorithms, such as making obvious errors and outputting
Markov Equivalence Classes (MECs). To address these challenge, we propose an
iterative method to assist researchers in manually constructing or modifying
DAGs by combining implied Conditional Independence (CI) testing with expert
knowledge. We introduce a novel measure of partial association for mixed data
to rank violations of implied CIs, enabling users to prioritize model
modifications while leveraging their expertise to decide the orientation of
edges when making these modifications. Empirical results show that this
approach achieves performance comparable to automated algorithms, assuming the
user can correctly orient edges in one out of three cases. Additionally, we
demonstrate that, in the absence of an expert, a Large Language Model can
potentially determine edge orientations.

\end{abstract}

@@

Despite the many automated algorithms proposed for constructing DAGs, their
adoption in applied research remains limited, with researchers often preferring
to manually construct DAGs based on their domain expertise \citep{Tennant2020,
Petersen2021}. We believe this preference persists due to several issues with
these algorithms when applied in practical settings with finite sample size:

\begin{enumerate}
	\item \textbf{Lack of Trust:} Although most algorithms are
		asymptotically consistent, their finite sample properties are
		not well understood \todo{Is there any citation for this}.
		Moreover, their output can vary significantly depending on the
		chosen algorithm and their hyperparameters. Combined with the
		lack of robust methods to test these algorithms' performance on
		a given dataset, it can be difficult to trust their outputs.
		Consequently, even when using these algorithms, users have to
		rely on post hoc manual modifications to the model based on model
		testing and their domain expertise.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} Multiple
		DAGs can be faithful to a given observational dataset. Thus,
		using only observational data, algorithms can only recover the
		MECs that can contain a combination of directed and undirected
		edges. However, most methods for downstream tasks, such as
		identification or causal effect estimation, assume knowledge of
		a fully oriented DAG. As a result, researchers have to manually
		orient these edges based on their expert knowledge before using
		them for analysis.
\end{enumerate}

@@

As a result of these issues, whether constructing the model manually or using
an algorithm, the user has to rely on their expertise to make modifications to
the model. One of the common ways to guide this post hoc modification is to use
implied CI based model testing \citep{Ankan2021, Ankan2023}. In this testing
procedure, users test whether the implied CIs of the DAG hold in the dataset or
not using statistical CI tests. Based on the results of these statistical
tests, we can accordingly make modifications to the current model to fix the
violations. However, this CI based testing also has a few drawbacks: 1) CIs are
implied only by missing edges in the model, so we can not test for already
existing edges. 2) In high sample size scenarios, p-values can suggest a
significant association between variables even if they have a very weak
relationship. 3) Number of implied CIs can be large making it difficult to
manually go through each of them; for example, in the moderately sized alarm
network with $ 37 $ node, there are $287$ implied CIs.

To tackle these issues with CI based testing, we take inspiration from
\emph{modification indices} in specification search. The modification indices
method tries to identify and rank changes in the model such that modifications
that improve the model fit the most are ranked higher.This ranking allows
researchers to focus first on the suggestions that can improve the model the
most and while integrating their expertise to choose the most optimal
modifications. Analogous to this, in this paper we develop a measure for
partial association for mixed data that allows us to rank the CI violations.
Additionally, this measure of conditional association also allows us tackle
other issues with CI based testing outlined in the previous paragraphs.

@@

\begin{figure}
\begin{subfigure}{0.5 \textwidth}
	\includegraphics[page=1]{figures.pdf}
\end{subfigure}
\begin{subfigure}{0.5 \textwidth}
	\includegraphics[page=2]{figures.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
	\includegraphics[page=3]{figures.pdf}
\end{subfigure}
